Climate-driven Phytoplankton community variability in the Cariaco Basin, Venezuela

### 2024-12-19 agujournaltemplate.tex

\title{Bottum-up drivers of phytoplankton community diversity in the Cariaco Basin, Venezuela}

@@

\begin{keypoints}
\item The phytoplankton community data identified by microscopy from the CARIACO Time Series between 1996 and 2016 shows two separate clusters characterized by high and low upwelling regimes respectively.
\item Despite the differing physical regimes and large shifts between functional groups, there is no significant difference in diversity metrics between the two clusters. Overall, diversity (i.e. Genus Richness) shows a positive linear relationship to Biomass, Productivity as well as Resource Use Efficiency.
\item A gradient forest analysis of the community data reveals that the strongest predictor of shifts within the community is the AMO index with a time lag of 5 months, which points to a strong impact of large scale climatic oscillations in the tropical coastal ecosystem. The next strongest predictor is nitrate concentration, with the MEI v2 index (lag of 4 months) and salinity following behind. 
\end{keypoints}

@@

\begin{abstract}
[ enter your Abstract here ]
\end{abstract}

@@

    Between late 1995 and early 1998, nutrient data were supplied by William Senior (Universidad de Oriente, Cumana, Venezuela) and from mid-1997 to the present by Dr. K. Fanning (University of South Florida, St. Petersburg, FL). Samples from eight cruises were analyzed by both laboratories. Data for $PO_4$ and $SiO_4$ were in strong agreement (within ±5\%), whereas the $NO_x$ and $NH_4$ measurements showed discrepancies. \cite{taylor_ecosystem_2012}. For the overview of the entire time-series, we used a merged dataset for all nutrients, where coincidental monthly measurements were averaged. For the analysis of Resource Use Efficiency, only the USF $NO_3$ measurements were used.

    For the identification of phytoplankton species by microscopy, 500 mL seawater samples were collected at the same depths in HDPE bottles and preserved with a 5\% (final concentration) formalin solution neutralized with sodium tetraborate. Quantitative analysis was performed at the Universidad de Oriente, Venezuela, following the Utermöhl method \cite{hasle1978inverted}, using 100 mL sedimentation chambers with a 48-hour settling period. Phytoplankton identification and enumeration by microscopy were restricted to microphytoplankton (cells $>$20 μm) \cite{mutshinda_environmental_2013}.
    %from Mutshinda et al. 2013
    To quality control the microscopy data, we assembled the taxonomic identifications made across all cruises, removed observations that were not identified to a species or genus level, corrected spelling variations, and grouped synonyms using taxonomic information in the World Register of Marine Species (\url{marinespecies.org}). After processing the data, there were 482 phytoplankton species observed over the course of the time series, which group into 223 phytoplankton genera.

    The datasets for Niskin bottle samples, CTD and phytoplankton taxonomy were retrieved from the Biological and Chemical Oceanography Data Management Office (BCO-DMO) \url{https://www.bco-dmo.org/project/2047} (accessed February, 2024).

    From the CARIACO Time Series data, we used only those variables, that cover the entire length of the time series without large gaps in coverage. Therefore, we did not include Zooplankton net tow and HPLC pigment data in our analysis.

@@

    \subsubsection{Data processing}
    % Niskin, CTD and Phytoplankton data processing
    Data was processed and analyzed using the R programming langauge (version 4.3.3) \cite{r_core_team_r_2024} using data pipelines created with the tidyverse packages \cite{wickham_welcome_2019}. All data processing and analysis code is publically available on Github \url{https://github.com/ben1post/cariaco-ms2}. %Note: Repo currently still set to private
    For all discrete depth measurements, the data were interpolated using the "oce" package \cite{kelley_oce_2023} with the adaptive "unesco" algorithm used by the U.S. National Oceanographic Data Center \cite{johnson2006world}. The interpolated values were depth integrated to 100 meters and in four discrete intervals between 0-25, 25-50, 50-75 and 75-100 meters for further analysis. For the phytoplankton microscopy data, the individual identified taxonomic units (Species or Genus) were interpolated, integrated, and then grouped by Genus via the sum of interpolated counts.

@@

    \subsubsection{Gradient Forest}
    To quantify the effects of bottom-up drivers in the Cariaco basin on shifts in the phytoplankton community, we employed a Gradient Forest analysis. Gradient Forest is a community-level extension of random forest analysis that combines regression trees calculated for indicator variables (here: monthly observed genus counts) to assess turnover in indicator variables across predictor variables\cite{pitcher_example_2012, large_critical_2015, tam_comparing_2017}. It calculates a goodness-of-fit statistic for each predictor variable and indicator variable, then averages these to determine the overall importance of each predictor. The regression tree split values also identify the ranges of predictors in which significant changes occur in the indicator variables, signaling a threshold for community turnover in the phytoplankton community. Cumulative plots of the predictive importance of each predictor variable can reveal thresholds for ecosystem drivers, as steep increases in cumulative importance signal a threshold for community turnover in the phytoplankton community \cite{tam_comparing_2017}. This analysis was conducted using monthly measurements integrated to 100 m depths for observed genera counts, ensuring the inclusion of all available data. The drivers/predictor variables considered in the analysis were the AMO index, MEI v.2 index, the -u 10m wind component, the 21 ° C isotherm, integrated temperature and salinity, as well as the concentration of nutrients ($NO_3$, $PO_4$ and $SiO_4$). The environmental variables from ocean time series tend to be strongly correlated, which the Gradient Forest methodology takes into account by implementing a conditional permutation of the correlated predictors \cite{ellis_gradient_2012}. To take into account the temporal lag in effect, in particular of climate indices, we included time lags up to 6 months of all variables in exploratory runs, and chose time lags that scored highest in the goodness-of-fit statistic for the final model run. % see figure in supplementary with initial runs TODO

\section{Draft Results}
Our study focused on phytoplankton microscopy data from the CARIACO time series, which consists of monthly samplings taken at multiple depths between November 1995 and January 2017. We compared these data with Niskin bottle nutrient, temperature, and salinity data, as well as climate data and indices, focusing on bottom-up drivers of community variability and diversity. In total, we used a complete set of 254 monthly data points, on which all variables were available. 
% Figure 1
The normalized yearly means for all environmental variables considered, as well as the logarithmic transformation of chlorophyll a and phytoplankton cell counts per functional group, are shown in Figure \ref{fig:zscore}. The patterns emerging from the -u 10m wind component, the temperature and the directly related 21 ° C isotherm clearly show the years of positive anomalies of wind (blue), upwelling (blue, i.e., shallower depth of the isotherm) and negative for temperature (red) within the top 100 m. These patterns are reflected in the mode of the Atlantic Multidecadal Oscillation (AMO), where a positive anomaly seems to correlate with weak upwelling regimes. The yearly mean nutrient concentrations have a maximum positive anomaly at the start of the time series, but show differing interannual oscillations. Annual mean anomalies of nitrate and phosphate seem to follow upwelling anomalies but remain relatively low between 2002 and 2014, with some more variability in mean phosphate concentration. Silicate does not follow a clear pattern, but shows an elevated mean concentration toward the later part of the time series (between 2008 and 2013), which could be driven by increased surface runoff \cite{lorenzoni_characterization_2015} and reduced depletion via diminished Diatom abundances.

% Figure 2
To further investigate the depth distribution and dynamics of chlorophyll biomass and community diversity, Figure \ref{fig:divts} shows the depth distributions over time, as well as across seasons. The bulk of the chlorophyll biomass ($\sim$ 80\%) was measured in the upper 50 meters of the water column, with approximately 15\% measured between 50 and 75 meters. Chlorophyll shows marked seasonality, with the highest biomass in the upper water column during the winter months (December–February, see Figure \ref{fig:divts} $b$). The yearly mean chlorophyll biomass shows the highest variability within the upper 25 meters, which during some years is below the concentration between 25 and 50 meters during the years 2000, 2005, 2007, 2008 and 2013 (see Figure \ref{fig:divts} $a$). In general, yearly mean anomalies within the upper 75 meters appear to be elevated during the period 1996 to 2004 and only recover in 2012. The year 2013 is an outlier, particularly for the upper 25 meters, as the missing monthly samplings from the winter month (January and February) skew the value. Nevertheless, the monthly measurements for the remainder of the year 2013 show a markedly lower biomass (65\% compared to the mean over the entire time series for these months), which recovers in 2014. Chlorophyll biomass below 50 meters depth shows a relatively stable distribution throughout the year and across the time series. In Fall, during the rainy season, mean chlorophyll biomass in the upper 25 meters is markedly reduced, while the 50–75 meters depth interval sees a slight increase, showing a shift in the chlorophyll maximum to lower depth under weak upwelling conditions.
The calculated genus diversity from the microscopy cell counts generally follows the dynamics of chlorophyll biomass in seasonality, with maximum values observed during the upwelling season in winter. Across the time series, we have maximum richness at the start of the time series (1996-1997), then a reduction in diversity across all depths, which only recovers past 2011. Scaling this genus diversity by the cell count abundances, i.e. the Shannon Index, shows a similar but less pronounced dynamic. Diversity remains high throughout 2000 and recovers more quickly, but shows marked reductions in annual means, particularly in the upper 25 meters in 2006, 2009 and 2016. The seasonal signal is almost lost (see Figure \ref{fig:divts} $f$), yet there is a consistent pattern of higher Shannon diversity at lower depths for all seasons except for Fall. 
Community evenness as signified by the Pielou index shows no seasonal signal and a very large variance within seasons (see Figure \ref{fig:divts} $h$). The interannual mean dynamics of evenness are remarkably stable and similar across all depth intervals, clustered around an intermediate 0.5 value. There are marked reductions in all depths towards 2000 and for all depths above 75 meters in 2006, 2009 and 2016. Throughout 2003 to 2013 evenness is almost consistently highest in the deepest interval below 75 meters, where the community seems to have been more evenly distributed.

>>>>

\title{Climate-driven Phytoplankton community variability in the Cariaco Basin, Venezuela}

@@

\begin{keypoints}
\item The CARIACO time series showed a highly variable phytoplankton community over the course of sampling between 1995 and 2017. Clustering of community data from microscopy cell counts reveals two separate clusters, defined by high- and low-upwelling conditions that match the AMO index. 
\item Despite the differing physical regimes and the large changes between the functional groups, there are no significant differences in genus richness and Shannon diversity between the two clusters, although community evenness was higher in low upwelling conditions.
\item A gradient forest analysis of community data reveals that the strongest predictor of shifts within the community is the AMO index with a time lag of 2 months, pointing to a strong impact of large-scale climatic oscillations in the tropical coastal ecosystem. The next strongest predictor is the MEI v2 index (lag of 4 months), with in-situ temperature and nitrate concentration following behind. 
\end{keypoints}

@@

\begin{abstract}
The structure of the phytoplankton community was identified using microscopy in the Cariaco basin, in Venezuela, between 1995 and 2017. Microscopy data are limited to larger cell sizes (> 20 µm), where the community showed a large change in the abundances of functional groups and shifts in diversity patterns. 
\end{abstract}

@@

    Between late 1995 and early 1998, nutrient data were supplied by William Senior (Universidad de Oriente, Cumana, Venezuela) and from mid-1997 to the present by Dr. K. Fanning (University of South Florida, St. Petersburg, FL). Samples from eight cruises were analyzed by both laboratories. Data for $PO_4$ and $SiO_4$ were in strong agreement (within ±5\%), whereas the $NO_x$ and $NH_4$ measurements showed discrepancies. \cite{taylor_ecosystem_2012}. For the overview of the entire time-series, we used a merged dataset for all nutrients, where coincidental monthly measurements were averaged.

    For the identification of phytoplankton species by microscopy, 500 mL seawater samples were collected at the same depths in HDPE bottles and preserved with a 5\% (final concentration) formalin solution neutralized with sodium tetraborate. Quantitative analysis was performed at the Universidad de Oriente, Venezuela, following the Utermöhl method \cite{hasle1978inverted}, using 100 mL sedimentation chambers with a 48-hour settling period. Phytoplankton identification and enumeration by microscopy were restricted to microphytoplankton (cells $>$20 μm) \cite{mutshinda_environmental_2013}.
    %from Mutshinda et al. 2013
    Quality control the microscopy data was performed by assembling the taxonomic identifications made across all cruises, removing observations that were not identified to species or genus level, correcting spelling variations, and grouping synonyms using taxonomic information in the World Register of Marine Species (\url{marinespecies.org}). After processing the data, 482 phytoplankton species were observed over the course of the time series, which were grouped into 223 phytoplankton genera.

    The datasets for Niskin bottle samples, CTD and phytoplankton taxonomy were retrieved from the Biological and Chemical Oceanography Data Management Office (BCO-DMO) \url{https://www.bco-dmo.org/project/2047} (accessed February, 2024). 
    From the CARIACO Time Series data, we used only those variables that cover the entire length of the time series without large gaps in coverage. Therefore, we did not include Zooplankton net tow and HPLC pigment data in our analysis.

@@

    \subsubsection{Data processing}
    % Niskin, CTD and Phytoplankton data processing
    Data were processed and analyzed using the R programming language (version 4.3.3) \cite{r_core_team_r_2024} using data pipelines created with tidyverse packages \cite{wickham_welcome_2019}. All data processing and analysis code is publically available on GitHub \url{https://github.com/ben1post/cariaco-ms2}. %Note: Repo currently still set to private
    For all discrete depth measurements, the data were interpolated using the "oce" package \cite{kelley_oce_2023} with the adaptive "unesco" algorithm defined by the U.S. National Oceanographic Data Center \cite{johnson2006world}. The interpolated values were depth integrated to 100 meters and in four discrete intervals between 0-25, 25-50, 50-75 and 75-100 meters for further analysis. For phytoplankton microscopy data, individual identified taxonomic units (Species or Genus) were interpolated, integrated, and then grouped by Genus via the sum of interpolated counts.

@@

    \subsubsection{Gradient Forest}
    To quantify the effects of bottom-up drivers in the Cariaco basin on shifts in the phytoplankton community, we employed a Gradient Forest analysis. Gradient Forest is a community-level extension of random forest analysis that combines regression trees calculated for indicator variables (here: monthly observed genus counts) to assess turnover in indicator variables across predictor variables \cite{pitcher_example_2012, large_critical_2015, tam_comparing_2017}. Gradient Forest calculates a goodness-of-fit statistic for each predictor variable and indicator variable, then averages these to determine the overall importance of each predictor. The regression tree split values also identify the ranges of predictors in which significant changes occur in the indicator variables, signaling a threshold for phytoplankton community turnover. Cumulative plots of the predictive importance of each predictor variable can reveal thresholds for ecosystem drivers, as steep increases in cumulative importance signal a threshold for community turnover in the phytoplankton community \cite{tam_comparing_2017}. This analysis was conducted using monthly measurements integrated to 100 m depths for observed genera counts, ensuring the inclusion of all available data. The drivers/predictor variables considered in the analysis were the AMO index, the MEI v.2 index, the -u 10m wind component, the 21 ° C isotherm, integrated temperature and salinity, as well as the concentration of nutrients ($NO_3$, $PO_4$ and $SiO_4$). The environmental variables from ocean time series tend to be strongly correlated, which the Gradient Forest methodology takes into account by implementing a conditional permutation of the correlated predictors \cite{ellis_gradient_2012}. To consider the temporal lag in effect, in particular of climate indices, we included time lags up to 6 months of all climate variables, as well as 3 months lags for in-situ measurements in exploratory runs, and chose time lags that scored highest in the goodness-of-fit statistic for the final model run (see Table \ref{app:table1} in the Appendix).

\section{Draft Results}
Our study focused on phytoplankton microscopy data from the CARIACO time series, which consists of monthly samplings taken at multiple depths between November 1995 and January 2017. We compared these data with Niskin bottle nutrient, temperature, and salinity data, as well as climate data and indices, focusing on bottom-up drivers of community variability and diversity. In total, we used a complete set of 254 monthly data points, on which all variables were available. 
% Figure 1
The normalized yearly means for all environmental variables considered, as well as the logarithmic transformation of chlorophyll a and phytoplankton cell counts per functional group, are shown in Figure \ref{fig:zscore}. The patterns emerging from the -u 10m wind component, the temperature and the directly related 21 ° C isotherm clearly show the years of positive anomalies of wind (blue), upwelling (blue, i.e., shallower depth of the isotherm) and negative for temperature (red) within the top 100 m. These patterns are reflected in the mode of the Atlantic Multidecadal Oscillation (AMO), where a positive anomaly seems to correlate with weak upwelling regimes. The yearly mean nutrient concentrations have a maximum positive anomaly at the start of the time series, but show differing inter-annual oscillations. Annual mean anomalies of nitrate and phosphate seem to follow upwelling anomalies but remain relatively low between 2002 and 2014, with some more variability in mean phosphate concentration. Silicate does not follow a clear pattern, but shows an elevated mean concentration toward the later part of the time series (between 2008 and 2013), which could be driven by increased surface runoff \cite{lorenzoni_characterization_2015} and reduced depletion via diminished Diatom abundances.

% Figure 2
To further investigate the depth distribution and dynamics of chlorophyll biomass and community diversity, Figure \ref{fig:divts} shows the depth distributions over time, as well as across seasons. The bulk of the chlorophyll biomass ($\sim$ 80\%) was measured in the upper 50 meters of the water column, with approximately 15\% measured between 50 and 75 meters. Chlorophyll shows marked seasonality, with the highest biomass in the upper water column during the winter months (December–February, see Figure \ref{fig:divts} $b$). The yearly mean chlorophyll biomass shows the highest variability within the upper 25 meters, which during some years is below the concentration between 25 and 50 meters during the years 2000, 2005, 2007, 2008 and 2013 (see Figure \ref{fig:divts} $a$). In general, yearly mean anomalies within the upper 75 meters appear to be elevated during the period 1996 to 2004 and only recover in 2012. The year 2013 is an outlier, particularly for the upper 25 meters, as the missing monthly samplings from the winter month (January and February) skew the value. Nevertheless, the monthly measurements for the remainder of the year 2013 show a markedly lower biomass (65\% of the mean over the entire time series for these months), which recovers in 2014. Chlorophyll biomass below 50 meters depth shows a relatively stable distribution throughout the year and across the time series. In Fall, during the rainy season, mean chlorophyll biomass in the upper 25 meters is markedly reduced, while the 50–75 meters depth interval sees a slight increase, showing a shift in the chlorophyll maximum to lower depth under weak upwelling conditions.
The calculated genus diversity from the microscopy cell counts generally follows the dynamics of chlorophyll biomass in seasonality, with maximum values observed during the upwelling season in winter. Across the time series, we have maximum richness at the start of the time series (1996-1997), then a reduction in diversity across all depths, which only recovers past 2011. Scaling this genus diversity by the cell count abundances, i.e. the Shannon Index, shows a similar but less pronounced dynamic. Diversity remains high throughout 2000 and recovers more quickly, but shows marked reductions in annual means, particularly in the upper 25 meters in 2006, 2009 and 2016. The seasonal signal is almost lost (see Figure \ref{fig:divts} $f$), yet there is a consistent pattern of higher Shannon diversity at lower depths for all seasons except for Fall. 
Community evenness as signified by the Pielou index shows no seasonal signal and a very large variance within seasons (see Figure \ref{fig:divts} $h$). The inter-annual mean dynamics of evenness are remarkably stable and similar across all depth intervals, clustered around an intermediate 0.5 value. There are marked reductions in all depths towards 2000 and for all depths above 75 meters in 2006, 2009 and 2016. Throughout 2003 to 2013 evenness is almost consistently highest in the deepest interval below 75 meters, where the community seems to have been more evenly distributed.
